---
title: "Data Frames"
author: "Kelsey Wickerham"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

# SET GLOBAL KNITR OPTIONS

knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10, 
                      fig.height = 8)


# LOAD PACKAGES

library(pander)
library(kableExtra)
library(dplyr)
library(ggplot2)
```

<br>
<br>

# Introduction

This code through will provide an overview to constructing data frames in R. Data frames are an 
essential structure used for data manipulation and analytics. It is made up of rows and columns,
and unlike matrices, it can contain different data types.

<br>

## Content Overview

In this code through we will specifically review;

1. The key features of a data frame

2. How to create a data frame from vectors

3. Accessing specific information stored in a data frame


<br>

## Why You Should Care

As previously mentioned, data frames are a fundamental structure in R. They offer versatility for
the manipulation and analysis of tabular data. As long as you are programming in R, you are going
to need to know what a data frame is, how to construct one, and how to access specific data.


<br>
<br>



# Step 1 - Key Features of a Data Frame

A data frame is made up of rows and columns. Columns represent the variable while rows represent 
what is observed. Because we typically name our columns in a data frame, it is easy to call upon 
our variables by their name when accessing specific data later on.

Unlike matrices, which can only contain one data type, a data frame can contain different types 
of data in the columns. This is important seeing as tabular data will often contain a combination 
of numeric, character, categorical, and logical data.

<br>
<br>

# Step 2 - Creating a Data Frame

We will be creating a data frame using information about players in my fantasy football team. 
This data frame will contain all four data types:

1. Characters - Player name
2. Categorical Data - Player position
3. Numeric - Total fantasy points scored this season, average score
4. Logical - Is the player eligible to play?

Now let's begin!

```{r}
#Step 1. Create the data frame using the data.frame function. For convention, we will name our 
#data frame "df".
#Step 2. Create the player vector and enter your player's names.
#Step 3. Create the position vector and enter the player positions.
#Step 4. Create the vector for total fantasy points scored and enter thenumeric data.
#Step 5. Create the vector for average points scored and enter the numeric data.
#Step 6. Create the vector identifying if a player is eligible to play. This is a T/F variable.

df <- data.frame(
  Player = c("Kyler Murray", "Breece Hall", "Puka Nacua", "Kyle Pitts", "Jake Moody"),
  Position = c("Quarterback", "Running Back", "Wide Receiver", "Tight End", "Kicker"),
  Total_Fantasy_Points = c(150.7, 68.7, 7.7, 38.0, 71.0),
  Average_Fantasy_Points = c(30.1, 13.7, 7.7, 7.6, 14.4),
  Eligible_To_Play = c(TRUE, TRUE, FALSE, TRUE, FALSE)
  
)

print(df)
```
Helpful tips when creating your data frame:

Column names cannot contain spaces. If a name has multipe words, such as "Total Fantasy Points", 
you should combine the name using underscores, periods, dashes, or "camel" spelling.

Additionally, make sure you use the equal sign when creating your vectors as opposed to an arrow.

<br>
<br>

# Step 3 - Accessing Information in a Data Frame

There are a variety of ways that you can access information in a data frame:

1. Accessing Columns by Name

To access a column by name, you can utilize the $ operator. Your code would look something like: 
   
   Data_Frame$Column_Name

For example, if I wanted to identify the names of my players, my code would simply be:
   
```{r}
df$Player
```
2. Accessing Rows and/or Columns by Position

You can also access information from a row and/or column based off its position in the table. 
This is achieved by utilizing brackets that contain the row and/or column number you are 
interested in.

   Data_Frame[Row Number, Column Number] - this will identify a very specific variable. 
   
For example, if I wanted to identify Kyler Murray's total fantasy points this season, I would 
write:

```{r}
df[1, 3]        #identifies the data in row 1, column 3
```

Perhaps I want to identify all of the data in row 1. In this case, I would write:


```{r}
df[1, ]          #identifies the data in row 1
```
You could also pull in multiple rows by writing:

   
```{r}
df[1:3, ]
```

And finally, if you wanted to identify specific columns - such at the player name and eligibility 
- you could write:
   
```{r}
df[, c(1, 5)]
```


<br> 
<br>

# Further Resources

Learn more about [package, technique, dataset] with the following:

<br>

* Resource I [Hyperlink Text](https://www.datacamp.com/tutorial/intro-data-frame-r)



<br>
<br>